Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-15 main.tex
unification nets

\newcommand{\unifstr}[1]{\mathcal{U}(#1)}

@@

In this paragraph, we associate each formula $A$ with its \bfit{formula tree}
$\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root.

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a \bfit{dualizer} $\dualizer$ of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}\lutz{use $\delta$ for the dualizer (or even better, make it a macro)}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

Let $\lambda$ is a linking on $\Gamma$ that has a dualizer.
The \bfit{unification structure} $\unifstr{\lambda}$
associated with $\lambda$ is the formula tree $\formtree{\form{\Gamma}}$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\single \exists x$ to $\single \forall y$ for every
dependency $(\single \exists x, \single \forall y)$.

@@

\begin{proof}
  We proceed by induction on the proof of $\sqn \Gamma$ in $\FOMLL$, making a
  case analysis on the bottommost rule instance:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\axr}{}{\sqn{a,\cneg a}}{}$ : the linking $\{a, \dual{a}\}$ is correct.
    \item $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ : the empty linking is correct.
    \item $\vliinf{\mixr}{}{\sqn{\Gamma,\Delta}}{\sqn\Gamma}{\sqn\Delta}$ :
	  By induction hypothesis, there is a correct linking on
	  $\Gamma$ and another one on $\Delta$, their union giving a
	  correct linking on $\Gamma, \Delta$.
    \item $\vlinf{\vlor}{}{\sqn{\Gamma,A\vlor B}}{\sqn{\Gamma,A,B}}$ : By
	  induction hypothesis, there is a correct linking on $\Gamma, A, B$,
	  and it is correct on $\Gamma, A \vlor B$ as well.
    \item $\vliinf{\vlan}{}{\sqn{\Gamma,A\vlan B, \Delta}}{\sqn{\Gamma,A}}{\sqn{B,\Delta}}$ :
          By induction hypothesis, there is a correct linking on $\Gamma, A$ and
	  another one on $B, \Delta$, their union giving a correct linking on
          $\Gamma,A \vlan B, \Delta$.
    \item $\vlinf{\exists}{}{\sqn{\Gamma,\exists x.A}}
	  {\sqn{\Gamma,A\ssubst{x}{t}}}$ : By induction hypothesis, there is a
	  correct linking $\lambda$ on $\Gamma, A\ssubst{x}{t}$.
	  For each atom in $\Gamma, A \ssubst{x}{t}$, there is a corresponding
	  atom in $\Gamma, \exists x.A$. There is therefore a linking $\lambda'$
          on $\Gamma, \exists x.A$ obtained from $\lambda$ via this
	  correspondence, and it is not difficult to check that $\lambda'$ is
	  correct as well.
    \item $\vlinf{\forall}{\text{($x$ not free in
	  $\Gamma$)}}{\sqn{\Gamma,\forall x.A}}{\sqn{\Gamma,A}}$ : By induction
          hypothesis, there is a correct linking on $\Gamma, A$, and it is easy
	  to see that it is a correct linking on $\Gamma, \forall x.A$ as well.
  \end{itemize}
  This allows us to define a translation from proofs in $\FOMLL$ to unification
  nets.
\end{proof}

\begin{thm}
  Any unification net can be obtained via the translation given in Theorem
  \ref{thm:MLL1->Unet}.
\end{thm}

>>>>

\newcommand{\unifstr}[1]{\mathcal{U}(#1)}
\newcommand{\sqntl}[1]{\left\lceil{#1}\right\rceil}

@@

In this paragraph, we associate each formula $A$ with its \bfit{formula tree}
$\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root. For a sequent
$\Gamma = A_1, \ldots, A_n$, we denote with $\formtree{\Gamma}$, the forest
formed by $\formtree{A_1}, \ldots, \formtree{A_n}$, i.e., the disjoint union of
$\formtree{A_i}$'s. The \bfit{roots} of $\formtree{\Gamma}$ are the roots of
${A_i}$'s

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the forest
$\formtree{\Gamma}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\formtree{\Gamma}$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a \bfit{dualizer} $\dualizer$ of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}\lutz{use $\delta$ for the dualizer (or even better, make it a macro)}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

Let $\lambda$ is a linking on $\Gamma$ that has a dualizer.
The \bfit{unification structure} $\unifstr{\lambda}$
associated with $\lambda$ is the forest $\formtree{\Gamma}$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\single \exists x$ to $\single \forall y$ for every
dependency $(\single \exists x, \single \forall y)$.

@@

\begin{proof}
  We proceed by induction on the proof of $\sqn \Gamma$ in $\FOMLL$, making a
  case analysis on the bottommost rule instance:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\axr}{}{\sqn{a,\cneg a}}{}$ : the linking $\{a, \dual{a}\}$ is correct.
    \item $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ : the empty linking is correct.
    \item $\vliinf{\mixr}{}{\sqn{\Gamma,\Delta}}{\sqn\Gamma}{\sqn\Delta}$ :
	  By induction hypothesis, there is a correct linking on
	  $\Gamma$ and another one on $\Delta$, their union giving a
	  correct linking on $\Gamma, \Delta$.
    \item $\vlinf{\vlor}{}{\sqn{\Gamma,A\vlor B}}{\sqn{\Gamma,A,B}}$ : By
	  induction hypothesis, there is a correct linking on $\Gamma, A, B$,
	  and it is correct on $\Gamma, A \vlor B$ as well.
    \item $\vliinf{\vlan}{}{\sqn{\Gamma,A\vlan B, \Delta}}{\sqn{\Gamma,A}}{\sqn{B,\Delta}}$ :
          By induction hypothesis, there is a correct linking on $\Gamma, A$ and
	  another one on $B, \Delta$, their union giving a correct linking on
          $\Gamma,A \vlan B, \Delta$.
    \item $\vlinf{\exists}{}{\sqn{\Gamma,\exists x.A}}
	  {\sqn{\Gamma,A\ssubst{x}{t}}}$ : By induction hypothesis, there is a
	  correct linking $\lambda$ on $\Gamma, A\ssubst{x}{t}$.
	  For each atom in $\Gamma, A \ssubst{x}{t}$, there is a corresponding
	  atom in $\Gamma, \exists x.A$. There is therefore a linking $\lambda'$
          on $\Gamma, \exists x.A$ obtained from $\lambda$ via this
	  correspondence, and it is not difficult to check that $\lambda'$ is
	  correct as well.
    \item $\vlinf{\forall}{\text{($x$ not free in
	  $\Gamma$)}}{\sqn{\Gamma,\forall x.A}}{\sqn{\Gamma,A}}$ : By induction
          hypothesis, there is a correct linking on $\Gamma, A$, and it is easy
	  to see that it is a correct linking on $\Gamma, \forall x.A$ as well.
  \end{itemize}
  This allows to define a translation $\sqntl\cdot$ from proofs in $\FOMLL$ to unification
  nets.
\end{proof}

\begin{thm}
  Any unification net can be obtained via the translation $\sqntl\cdot$ given in Theorem
  \ref{thm:MLL1->Unet}.
\end{thm}
Let $\lambda$ be a unification net on $\Gamma$. We proceed by induction on the
number of connectives in $\Gamma$. In the base case, $\Gamma$ is of the form
\begin{align*}
p_1(t_{11}, \ldots, t_{1n_1}), \dual{p_1}(t_{11}, \ldots, t_{1n_1}), \ldots, \\
p_k(t_{k1}, \ldots, t_{kn_k}), \dual{p_k}(t_{k1}, \ldots, t_{kn_k}),
\underbrace{\ttt, \ldots, \ttt}_\text{$m$ times}
\end{align*} and
$\lambda$ is the linking $\set{(a_1, \cneg{a_1}), \ldots, (a_k, \cneg{a_k})}$,
where $a_i = p_i(t_{i1}, \ldots, t_{in_i})$, which equals to $\sqntl\Pi$, where
$\Pi$ is the proof consisting of $m$ instances of the $\ttt$ rule, $n$ instances
$\vlinf{\ax}{}{\sqn a_i, \cneg{a_i}}{}$ of the $\ax$ rule, and followed by
$m+k-1$ instances of the $\mix$ rule.

Now we consider the inductive cases:
\begin{itemize}
  \item $\Gamma = \Delta, A \vlor B$:
  \item $\Gamma = \Delta \forall x.A$:
\end{itemize}
